# Reproducible Research Assignment 1
Muralidhar Areti  
Jul 17, 2015

## Introduction
This R Markdown file contains the code and results of the work required for submission for the first peer assessment. 

### Loading and preprocessing the data
The first step of the assignment is to load the dataset. 
```{r}
suppressPackageStartupMessages(library(dplyr))
library(lubridate)
library(ggplot2)

activity_data = read.csv("activity.csv"
             , na.strings="NA"
             )
```

During my initial loads I like to print the dimensions, structure, summary and top few rows to get an idea of the data I am working with. 
```{r}
print(dim(activity_data))
print(str(activity_data))
print(summary(activity_data))
print(head(activity_data))
```

### What is the mean total number of steps taken per day?
```{r}
steps_per_day = activity_data %>%
    group_by(date) %>%
    summarize(sumSteps = sum(steps))

steps_per_day

# The "traditional" way to calculate the mean in R
ms = mean(steps_per_day$sumSteps, na.rm = TRUE)
ms

# The "traditional" way to calculate the median in R
ms2 = median(steps_per_day$sumSteps, na.rm = TRUE)
ms2

# Calculate the mean and median using dplyr
summarize(steps_per_day, mean_steps = mean(sumSteps, na.rm = TRUE))
summarize(steps_per_day, median_steps = median(sumSteps, na.rm = TRUE))

steps_per_day %>%
    ggplot(aes(x=factor(date), y=sumSteps)) +
    geom_bar(fill="steelblue", stat="identity", alpha=0.8) +
    labs(x="Date") +
    labs(y="Number of Steps") + 
    labs(title="Steps per Day") +
    geom_hline(aes(yintercept = ms)) + 
    geom_text(
        aes(0,ms, label=paste("mean =", round(ms,2)))
        , vjust=-0.5
        , hjust=-0.1
        , size=4
        ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle=90, size=8))
```


### What is the average daily activity pattern?
```{r}
activity_data %>%
    ggplot(aes(x=interval, y=steps) + 
    geom_line() +
    theme_bw()
```

### Imputing missing values


### Are there differences in activity patterns between weekdays and weekends?



