---
title: "Untitled"
output: 
    html_document: 
        keep_me: true
---
Muralidhar Areti  
July 24, 2015

## Introduction
This R Markdown file contains the code and results of the work required for submission for the first peer assessment. 

### Loading and preprocessing the data
The first step of the assignment is to load the dataset. 
```{r 1.1 - load data}
suppressPackageStartupMessages(library(dplyr))
library(lubridate)
library(ggplot2)

activity_data = read.csv("activity.csv"
             , na.strings="NA"
             )
```

```{r 1.2 - test data, results='hide'}
print(dim(activity_data))
print(str(activity_data))
print(summary(activity_data))
print(head(activity_data))
```

```{r 2.1 - steps per day, results='hide'}
steps_per_day = activity_data %>%
    group_by(date) %>%
    summarize(sumSteps = sum(steps, na.rm = TRUE))

steps_per_day
```

There are two ways to determine the mean and the median, the traditional way: 
```{r 2.1 - mean and median steps per day, results='hide'}
ms1 = mean(steps_per_day$sumSteps, na.rm = TRUE)
ms2 = median(steps_per_day$sumSteps, na.rm = TRUE)
```

And the by using dplyr:
```{r 2.1 - mean and median steps per day dplyr, results='hide'}
summarize(steps_per_day, mean_steps = mean(sumSteps, na.rm = TRUE))
summarize(steps_per_day, median_steps = median(sumSteps, na.rm = TRUE))
```

Below is the plot of the histogram showing how many steps were measured per day:
```{r 2.2 - histogram of steps per day}
steps_per_day %>%
    ggplot(aes(x=factor(date), y=sumSteps)) +
    geom_bar(fill="steelblue", stat="identity", alpha=0.8) +
    labs(x="Date") +
    labs(y="Number of Steps") + 
    labs(title="Steps per Day") +
    geom_hline(aes(yintercept = ms1)) + 
    geom_text(
        aes(0,ms1, label=paste("mean =", round(ms1,2)))
        , vjust=-0.5
        , hjust=-0.1
        , size=4
        ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle=90, size=8))
```

### What is the average daily activity pattern?
First we must perform some data management tasks to ensure we have the right details required for plotting. 
```{r 3.1 - daily activity pattern}
steps_per_interval = activity_data %>%
    group_by(interval) %>%
    summarize(
        sumSteps = sum(steps, na.rm = TRUE)
        , meanSteps = mean(steps, na.rm = TRUE)
        )
# steps_per_interval

ms3 = mean(steps_per_interval$meanSteps, na.rm = TRUE)
ms4 = steps_per_interval %>%
    filter(meanSteps == max(meanSteps)) %>%
    select(interval)

steps_per_interval %>%
    ggplot(aes(x=interval, y=meanSteps)) + 
    geom_line() +
    geom_hline(aes(yintercept = ms3)) + 
    geom_vline(xintercept = as.numeric(ms4), linetype="dashed") +
    geom_text(
        aes(as.numeric(ms4),0, label=paste("max =", round(as.numeric(ms4),0)))
        , vjust=0
        , hjust=-0.1
        , size=4
        ) +
    theme_bw()
```


### Imputing missing values
```{r 4.1 - checking for NAs per day}
na_per_day = activity_data %>%
    group_by(date) %>%
    summarize(
        sumSteps = sum(steps, na.rm=TRUE)
        , numRecords = n()
        , numNa = sum(is.na(steps))
        ) %>%
    filter(numNa > 0)

na_per_day
```

```{r 4.1 - summarize NAs for prose, echo=FALSE}
total_na = na_per_day %>%
    summarize(sumNa = sum(numNa)) %>%
    select(sumNa)
```

The results above shows `r nrow(na_per_day) ` reccords containing 1 or more NA values. The total number of NA values in the dataset is `r as.numeric(total_na)` (same number as shown in the summary). 

```{r 4.3 - Impute data, results='hide'}
average_steps = mean(activity_data$steps, na.rm=TRUE)
average_steps

imputed_data = activity_data %>%
    mutate(imputed_steps = ifelse(is.na(steps), average_steps, steps))

imputed_steps_per_day = imputed_data %>%
    group_by(date) %>%
    summarize(
        sumSteps = sum(steps, na.rm = TRUE)
        , imputed_sumSteps = sum(imputed_steps, na.rm=TRUE)
        )
```

```{r 4.3 Graphing the imputed data}
ms5 = mean(imputed_steps_per_day$imputed_sumSteps, na.rm = TRUE) 
ms6 = median(imputed_steps_per_day$imputed_sumSteps, na.rm = TRUE)

imputed_steps_per_day %>%
    ggplot(aes(x=factor(date), y=imputed_sumSteps)) +
    geom_bar(fill="slateblue", stat="identity", alpha=0.8) +
    labs(x="Date") +
    labs(y="Number of Steps") + 
    labs(title="Steps per Day") +
    geom_hline(aes(yintercept = ms5)) + 
    geom_text(
        aes(0,ms5, label=paste("mean =", round(ms5,2)))
        , vjust=-0.5
        , hjust=-0.1
        , size=4
        ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle=90, size=8))
```

### Are there differences in activity patterns between weekdays and weekends?
First lets check to see if we can determine which days are weekdays:
```{r 5.1 find and compare weekday vs weekend}
weekday_dataset = activity_data %>%
   mutate(
        date = as.Date(date)
        , weekdayName = weekdays(date)
        , weekendFlag = ifelse(
            weekdayName == "Sunday" | weekdayName == "Saturday"
            , "Weekend"
            , "Weekday"
            )
        ) %>%
    group_by(weekendFlag, interval) %>%
    summarize(meanSteps = mean(steps, na.rm=TRUE)) 
```

Now that the dataset has been set we can create our desired plot. 
```{r 5.1 Plot the weekday and weekend steps}
weekday_dataset %>%
    ggplot(aes(x=interval, y=meanSteps)) + 
    geom_line() +
    facet_grid(weekendFlag ~ .) +
    theme_bw()
```